---
title: "Final Presentation"
author: "Nathan Dhanani"
subtitle: "5/08/25"
format:
  revealjs:
    scrollable: true
    slide-number: true
    show-slide-number: all
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction & Methodology

**Research Question:** Does greater access to financial institutions correlate with higher national wealth, as measured by GDP per capita?

**Data Overview:**

-   120 countries

-   Variables sourced from World Bank/Wolrdometer

-   Countries grouped by median GDP per capita:

    -   *High GDP group*: Above median

    -   *Low GDP group*: Below median

**Main Tool:**

-   Permutation Test

-   Simulates the null hypothesis: no relationship between GDP and account access

-   1,000 iterations with random shuffling

## Data Preparation

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
library(tidyverse)
library(readxl)

GDP_analysis <- read_excel("GDP_analysis.xlsx", sheet = "Data", skip = 1)
colnames(GDP_analysis) <- c(
  "Country", 
  "GDP_Per_Capita", 
  "Account_15plus", 
  "Saved_15plus", 
  "Borrowed_Formal_15plus", 
  "Account_Female", 
  "Account_Male", 
  "Owns_Credit_Card")

DP_analysis <- GDP_analysis |> 
  mutate(across(GDP_Per_Capita:Owns_Credit_Card, as.numeric))
  
median_gdp <- median(GDP_analysis$GDP_Per_Capita, na.rm = TRUE)

GDP_analysis <- GDP_analysis |>  
  mutate(GDP_Group = ifelse(GDP_Per_Capita >= median_gdp, "High", "Low"))
```

## Box plot Visualization

```{r, boxplot-access-account, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE, fig.width=8, fig.height=5, out.width='800px', fig.align='center'}
GDP_analysis |>
  ggplot(aes( x = GDP_Group, y = Account_15plus )) +
  geom_boxplot()
```

## Permutation Test

**Null hypothesis:**

There is no difference in the degree of access to financial institutions between countries with higher and lower GDP per capita.

**Alternate Hypothesis:**

Countries with a higher GDP per capita have greater access to financial institutions than those with a lower GDP per capita.

```{r echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
perm_data <- function(rep, data) {
  data |>
    select(GDP_Group, Account_15plus) |>
    mutate(Account_15plus_perm = sample(Account_15plus, replace = FALSE)) %>%
    group_by(GDP_Group) |>
    summarize(
      obs_ave = mean(Account_15plus, na.rm = TRUE),
      obs_med = median(Account_15plus, na.rm = TRUE),
      perm_ave = mean(Account_15plus_perm, na.rm = TRUE),
      perm_med = median(Account_15plus_perm, na.rm = TRUE),
      .groups = "drop"
    ) |>
    summarize(
      obs_ave_diff = diff(obs_ave),
      obs_med_diff = diff(obs_med),
      perm_ave_diff = diff(perm_ave),
      perm_med_diff = diff(perm_med),
      rep = rep
    )
}

perm_results <- map_df(1:1000, perm_data, data = GDP_analysis)

head(perm_results)


```

## Permutation test results

```{r echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
ggplot(perm_results, aes(x = perm_ave_diff)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  geom_vline(aes(xintercept = mean(obs_ave_diff)), color = "red") +
  labs(
    title = "Permutation Test: Difference in Means of Account Ownership (15+)",
    x = "Difference in Means (Low GDP - High GDP)",
    y = "Count"
  ) +
  theme_minimal()

```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
ggplot(perm_results, aes(x = perm_med_diff)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  geom_vline(aes(xintercept = mean(obs_med_diff)), color = "blue",) +
  labs(
    title = "Permutation Test: Difference in Medians of Account Ownership (15+)",
    x = "Difference in Medians (Low GDP - High GDP)",
    y = "Count"
  ) +
  theme_minimal()
```
