---
title: "NFL Salaries by Position"
description: |
 [Tidy Tuesday source](https://github.com/rfordatascience/tidytuesday/tree/main/data/2018/2018-04-09)
 [Raw data source](https://www.spotrac.com/)
author: Nathan Dhanani
date: Febuary 19, 2025
format: html
---

Introduction:

To explore how player compensation varies across different roles in the NFL, I analyzed a dataset containing salary data by position. Initially, the data was in wide format, with each column representing a different player position. To better visualize and compare the salary distributions across these positions, I reshaped the dataset into a long format using a pivot operation. This allowed me to create a single box plot chart displaying the spread, median, and variability of salaries for each position. The resulting visualization reveals clear disparities—most notably, the consistently higher salaries earned by quarterbacks, while positions like special teamer and safeties tend to fall on the lower end of the pay scale. This plot offers an accessible summary of the economic hierarchy within professional football.

Exploratory Analysis:

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
library(readxl)
library(dplyr)

# Load the data
nfl_salary <- read_excel("nfl_salary.xlsx")

# Show the first few rows of the full dataset with all original variables
head(nfl_salary, 10)
```

Variables:

-   year - The year the salary data was recorded

-   Corner back - Salary of the top-paid cornerbacks that year (USD)

-   Defensive Lineman - Salary of the top-paid defensive linemen (USD)

-   Linebacker - Salary of the top-paid linebackers (USD)

-   Offensive Lineman - Salary of the top-paid offensive linemen (USD)

-   Quarterback - Salary of the top-paid quarterbacks (USD)

-   Running Back - Salary of the top-paid running backs (USD)

-   Safety - Salary of the top-paid safeties (USD)

-   Special Teamer - Salary of the top-paid special teams players (USD)

-   Tight End - Salary of the top-paid tight ends (USD)

-   Wide Receiver - Salary of the top-paid wide receivers (USD)

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE, fig.alt="This line plot shows average salaries for quarterbacks, running backs, and wide receivers from 2011 to 2020."}

# Load necessary libraries for data manipulation and visualization

library(tidyverse)
library(readxl)

# Read the Excel file containing NFL salary data

file_path <- "nfl_salary.xlsx"
nfl_salary <- read_excel(file_path)

# Reshape the data from wide to long format
# Keeps 'year' as is, and stacks the salary data by position into two columns: 'position' and 'salary'

nfl_long <- nfl_salary |>
  pivot_longer(
    cols = -year,
    names_to = "position",
    values_to = "salary"
  )

# Create a boxplot to visualize salary distribution by NFL position

ggplot(nfl_long, aes(x = position, y = salary)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(title = "NFL Salaries by Position", x = "Position", y = "Salary") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

This boxplot shows the distribution of NFL player salaries by position. Each box represents the range of salaries for a given position across multiple players, highlighting the median, quartiles, and potential outliers. It allows for easy comparison of pay differences between positions, such as the typically higher salaries of quarterbacks compared to special teamers.

Citations:

TidyTuesday source: Mock, T., & TidyTuesday Project. (2018, April 9). *NFL salary data*. TidyTuesday. Retrieved from <https://github.com/rfordatascience/tidytuesday/blob/main/data/2018/2018-04-09/nfl_salary.xlsx>

Original data source: Original data from Spotrac. (n.d.). *NFL player contracts*. Spotrac. Retrieved from https://www.spotrac.com/nfl/rankings/

Additional citation context:

-   Collector: The dataset was compiled and shared by Thomas Mock as part of the weekly TidyTuesday project.

-   Purpose: The dataset is intended for data visualization and wrangling practice in the R programming language.

-   Original source: The salary data was scraped from Spotrac, a site that tracks sports contracts and salaries.

-   Why collected: This dataset was curated to encourage open data analysis, reproducible research, and collaborative learning among the R community.
