---
title: "SQL"
description: |
 [Stanford Open Policing Project Data source ](https://openpolicing.stanford.edu/)
author: Nathan Dhanani
date: April 16, 2025
format: html
---

Introduction:

In this project, I investigate racial disparities in police search behavior during traffic stops across six U.S. cities: Nashville, Charlotte, Saint Paul, New Orleans, and Austin,  Using data from the Stanford Open Policing Project, I analyze whether Black and White drivers experience different search rates, and how those rates have changed over time. My approach involves querying each city’s data using SQL to summarize annual stop counts, search counts, and calculated search rates by race. I then compare these trends both within and across cities to identify patterns of disparity or improvement over time.


Exploratory analysis:
```{r echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
library(DBI)
library(RMariaDB)

# Establish a secure connection to the 'traffic' database using environmental credentials
con_traffic <- DBI::dbConnect(
  RMariaDB::MariaDB(),
  dbname = "traffic",
  host = Sys.getenv("TRAFFIC_HOST"),
  user = Sys.getenv("TRAFFIC_USER"),
  password = Sys.getenv("TRAFFIC_PWD")
)

# List all available tables in the connected database
DBI::dbListTables(con_traffic)

```

```{sql, connection=con_traffic, output.var="nashville_df"}
# Query Nashville traffic stop data to calculate search rates for Black and White drivers by year

SELECT 
  'Nashville' AS city,
  subject_race AS race,
  YEAR(date) AS year,
  COUNT(*) AS total_stops,
  SUM(search_conducted) AS searches,
  ROUND(SUM(search_conducted) * 1.0 / COUNT(*), 3) AS search_rate
FROM tn_nashville_2020_04_01
WHERE 
  subject_race IN ('black', 'white') AND
  date IS NOT NULL AND
  search_conducted IS NOT NULL
GROUP BY subject_race, YEAR(date)
ORDER BY subject_race, YEAR(date);
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
#| fig-alt: "This chart shows two vertical boxplots, one for black drivers and one for white drivers in Nashville. The x-axis shows the driver’s race, and the y-axis shows the search rate, which means how often drivers were searched during traffic stops. Each box shows the range of search rates across multiple years. The line in the middle of the box is the median (typical) search rate, and the height of the box shows how much the rates vary. We see that the box representing black individuals is much higher that white individual which shows that, on average, black drivers are searched more frequently than white drivers"

library(ggplot2)
library(dplyr)

# This code creates a boxplot comparing search rates for Black and White drivers in Nashville.
# The search rate is calculated as the number of searches divided by the total number of stops, and is visualized as a distribution for each racial group across all years combined.


nashville_df |>
  mutate(search_rate = searches / total_stops) |>
  ggplot(aes(x = race, y = search_rate, fill = race)) +
  geom_boxplot() +
  labs(
    title = "Overall Search Rate by Race in Nashville",
    x = "Driver Race",
    y = "Search Rate",
    fill = "Driver Race") +
  theme_minimal()
```
This graph shows a Boxplot that plots the distribution of search rates for Black and White drivers in Nashville. Each box represents the total number of search rates  within each racial group. The higher median for Black drivers suggests they were searched more frequently across all the years we have data forthan White drivers.


```{sql, connection=con_traffic, output.var="charlotte_df" }

# Query Charlotte traffic stop data to calculate search rates for Black and White drivers by year

SELECT 
  'Charlotte' AS city,
  subject_race AS race,
  YEAR(date) AS year,
  COUNT(*) AS total_stops,
  SUM(search_conducted) AS searches,
  ROUND(SUM(search_conducted) * 1.0 / COUNT(*), 3) AS search_rate
FROM tn_nashville_2020_04_01
WHERE 
  subject_race IN ('black', 'white') AND
  date IS NOT NULL AND
  search_conducted IS NOT NULL
GROUP BY subject_race, YEAR(date)
ORDER BY subject_race, YEAR(date);
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
#| fig-alt: "This chart shows two vertical boxplots, one for black drivers and one for white drivers in Charlotte The x-axis shows the driver’s race, and the y-axis shows the search rate, which means how often drivers were searched during traffic stops. Each box shows the range of search rates across multiple years. The line in the middle of the box is the median (typical) search rate, and the height of the box shows how much the rates vary. We see that the box representing black individuals is much higher that white individual which shows that, on average, black drivers are searched more frequently than white drivers"

# This code creates a boxplot comparing search rates for Black and White drivers in Charlotte
# The search rate is calculated as the number of searches divided by the total number of stops, and is visualized as a distribution for each racial group across all years combined.

charlotte_df |>
  mutate(search_rate = searches / total_stops) |>
  ggplot(aes(x = race, y = search_rate, fill = race)) +
  geom_boxplot() +
  labs(
    title = "Overall Search Rate by Race in Charlotte",
    x = "Driver Race",
    y = "Search Rate",
    fill = "Driver Race") +
  theme_minimal()
```

This graph shows a Boxplot that plots the distribution of search rates for Black and White drivers in Charlotte Each box represents the total number of search rates  within each racial group. The higher median for Black drivers suggests they were searched more frequently across all the years we have data forthan White drivers.


```{sql, connection=con_traffic, output.var="saint_paul_df"}

# Query Saint Paul traffic stop data to calculate search rates for Black and White drivers by year

SELECT 
  'Saint Paul' AS city,
  subject_race AS race,
  YEAR(date) AS year,
  COUNT(*) AS total_stops,
  SUM(search_conducted) AS searches,
  ROUND(SUM(search_conducted) * 1.0 / COUNT(*), 3) AS search_rate
FROM tn_nashville_2020_04_01
WHERE 
  subject_race IN ('black', 'white') AND
  date IS NOT NULL AND
  search_conducted IS NOT NULL
GROUP BY subject_race, YEAR(date)
ORDER BY subject_race, YEAR(date);
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
#| fig-alt: "This chart shows two vertical boxplots, one for black drivers and one for white drivers in Saint Paul The x-axis shows the driver’s race, and the y-axis shows the search rate, which means how often drivers were searched during traffic stops. Each box shows the range of search rates across multiple years. The line in the middle of the box is the median (typical) search rate, and the height of the box shows how much the rates vary. We see that the box representing black individuals is much higher that white individual which shows that, on average, black drivers are searched more frequently than white drivers"

# This code creates a boxplot comparing search rates for Black and White drivers in Saint Paul
# The search rate is calculated as the number of searches divided by the total number of stops, and is visualized as a distribution for each racial group across all years combined.

saint_paul_df |>
  mutate(search_rate = searches / total_stops) |>
  ggplot(aes(x = race, y = search_rate, fill = race)) +
  geom_boxplot() +
  labs(
    title = "Overall Search Rate by Race in Saint Paul",
    x = "Driver Race",
    y = "Search Rate",
    fill = "Driver Race") +
  theme_minimal()
```

This graph shows a Boxplot that plots the distribution of search rates for Black and White drivers in Saint Paul Each box represents the total number of search rates  within each racial group. The higher median for Black drivers suggests they were searched more frequently across all the years we have data forthan White drivers.


```{sql, connection=con_traffic, output.var="new_orleans_df"}

# Query New Orleans traffic stop data to calculate search rates for Black and White drivers by year

SELECT 
  'New Orleans' AS city,
  subject_race AS race,
  YEAR(date) AS year,
  COUNT(*) AS total_stops,
  SUM(search_conducted) AS searches,
  ROUND(SUM(search_conducted) * 1.0 / COUNT(*), 3) AS search_rate
FROM tn_nashville_2020_04_01
WHERE 
  subject_race IN ('black', 'white') AND
  date IS NOT NULL AND
  search_conducted IS NOT NULL
GROUP BY subject_race, YEAR(date)
ORDER BY subject_race, YEAR(date);
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
#| fig-alt: "This chart shows two vertical boxplots, one for black drivers and one for white drivers in New Orleans The x-axis shows the driver’s race, and the y-axis shows the search rate, which means how often drivers were searched during traffic stops. Each box shows the range of search rates across multiple years. The line in the middle of the box is the median (typical) search rate, and the height of the box shows how much the rates vary. We see that the box representing black individuals is much higher that white individual which shows that, on average, black drivers are searched more frequently than white drivers"

# This code creates a boxplot comparing search rates for Black and White drivers in New Orleans
# The search rate is calculated as the number of searches divided by the total number of stops, and is visualized as a distribution for each racial group across all years combined.

new_orleans_df |>
  mutate(search_rate = searches / total_stops) |>
  ggplot(aes(x = race, y = search_rate, fill = race)) +
  geom_boxplot() +
  labs(
    title = "Overall Search Rate by Race in New Orleans",
    x = "Driver Race",
    y = "Search Rate",
    fill = "Driver Race") +
  theme_minimal()
```

This graph shows a Boxplot that plots the distribution of search rates for Black and White drivers in New Orleans Each box represents the total number of search rates  within each racial group. The higher median for Black drivers suggests they were searched more frequently across all the years we have data forthan White drivers.


```{sql, connection=con_traffic, output.var="austin_df"}

# Query Austin traffic stop data to calculate search rates for Black and White drivers by year

SELECT 
  'Austin' AS city,
  subject_race AS race,
  YEAR(date) AS year,
  COUNT(*) AS total_stops,
  SUM(search_conducted) AS searches,
  ROUND(SUM(search_conducted) * 1.0 / COUNT(*), 3) AS search_rate
FROM tn_nashville_2020_04_01
WHERE 
  subject_race IN ('black', 'white') AND
  date IS NOT NULL AND
  search_conducted IS NOT NULL
GROUP BY subject_race, YEAR(date)
ORDER BY subject_race, YEAR(date);
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
#| fig-alt: "This chart shows two vertical boxplots, one for black drivers and one for white drivers in Austin The x-axis shows the driver’s race, and the y-axis shows the search rate, which means how often drivers were searched during traffic stops. Each box shows the range of search rates across multiple years. The line in the middle of the box is the median (typical) search rate, and the height of the box shows how much the rates vary. We see that the box representing black individuals is much higher that white individual which shows that, on average, black drivers are searched more frequently than white drivers"

# This code creates a boxplot comparing search rates for Black and White drivers in Austin
# The search rate is calculated as the number of searches divided by the total number of stops, and is visualized as a distribution for each racial group across all years combined.

austin_df |>
  mutate(search_rate = searches / total_stops) |>
  ggplot(aes(x = race, y = search_rate, fill = race)) +
  geom_boxplot() +
  labs(
    title = "Overall Search Rate by Race in Austin",
    x = "Driver Race",
    y = "Search Rate",
    fill = "Driver Race") +
  theme_minimal()
```

This graph shows a Boxplot that plots the distribution of search rates for Black and White drivers in Austin Each box represents the total number of search rates  within each racial group. The higher median for Black drivers suggests they were searched more frequently across all the years we have data forthan White drivers.


```{sql, connection=con_traffic, output.var="df"}

#This query extracts and combines stop and search counts by race and year for Black and White drivers in Nashville, Charlotte, Saint Paul, New Orleans, and Austin.

SELECT 'Nashville' AS city, subject_race AS race, YEAR(date) AS year,
       COUNT(*) AS total_stops,
       SUM(CASE WHEN search_conducted = TRUE THEN 1 ELSE 0 END) AS searches
FROM tn_nashville_2020_04_01
WHERE subject_race IN ('black', 'white')
GROUP BY subject_race, YEAR(date)

UNION ALL

SELECT 'Charlotte' AS city, subject_race AS race, YEAR(date) AS year,
       COUNT(*) AS total_stops,
       SUM(CASE WHEN search_conducted = TRUE THEN 1 ELSE 0 END) AS searches
FROM nc_charlotte_2020_04_01
WHERE subject_race IN ('black', 'white')
GROUP BY subject_race, YEAR(date)

UNION ALL

SELECT 'Saint Paul' AS city, subject_race AS race, YEAR(date) AS year,
       COUNT(*) AS total_stops,
       SUM(CASE WHEN search_conducted = TRUE THEN 1 ELSE 0 END) AS searches
FROM mn_saint_paul_2020_04_01
WHERE subject_race IN ('black', 'white')
GROUP BY subject_race, YEAR(date)

UNION ALL

SELECT 'New Orleans' AS city, subject_race AS race, YEAR(date) AS year,
       COUNT(*) AS total_stops,
       SUM(CASE WHEN search_conducted = TRUE THEN 1 ELSE 0 END) AS searches
FROM la_new_orleans_2020_04_01
WHERE subject_race IN ('black', 'white')
GROUP BY subject_race, YEAR(date)

UNION ALL

SELECT 'Austin' AS city, subject_race AS race, YEAR(date) AS year,
       COUNT(*) AS total_stops,
       SUM(CASE WHEN search_conducted = TRUE THEN 1 ELSE 0 END) AS searches
FROM tx_austin_2020_04_01
WHERE subject_race IN ('black', 'white')
GROUP BY subject_race, YEAR(date);
```

```{r echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
#| fig-alt: "This chart shows a series of small line graphs, one for each city. Each small graph has two colored lines — one for Black drivers and one for White drivers. The x-axis shows the year, and the y-axis shows the search rate, which is the percentage of traffic stops that led to a search. The lines show how search rates changed over time for each race within each city. We see that the black driver line is consistently higher than the white driver line meaning black people are searched more often."

# Create a line plot showing the proportion of stops that resulted in a search by race and city across years, faceted by city

library(ggplot2)
library(dplyr)

df |>
  mutate(year = as.numeric(year)) |>
  ggplot(aes(x = year, y = searches / total_stops, color = race)) +
  geom_line(size = 1.2) +
  facet_wrap(~ city) +
  labs(title = "Search Rate Over Time by Race and City",
       x = "Year", y = "Search Rate",
       color = "Driver Race") +
  theme_minimal()
```

This plot displays search rates over time for Black and White drivers in each of the six cities included in the analysis seperated into 6 graphs per city. Each line represents the percentage of stops that resulted in a search for each racial group, by year. Across all cities, Black drivers consistently experienced higher search rates than White drivers. The gap between the two groups is visible in most cities and remains relatively stable over time, suggesting a persistent racial disparity in search practices.



```{r echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
#| fig-alt: "This chart shows side-by-side bars for Black and White drivers in each of five cities. The x-axis shows the cities, and the y-axis shows the total number of traffic stops. Each city has two bars next to each other—one for each race—so you can directly compare how often each group was stopped in that city. we actually don't see a trend in which race gets stopped more so we know that Black drivers are searched at a hgiher rate not because they are stopped more but likely for other reasons such as racial bias."

# This code creates a bar chart showing how many traffic stops occurred for Black and White drivers in each of five cities.

df |>
  group_by(city, race) |>
  summarise(total_stops = sum(total_stops)) |>
  ggplot(aes(x = city, y = total_stops, fill = race)) +
  geom_col(position = "dodge") +
  labs(title = "Total Stops by Race and City",
       x = "City", y = "Total Stops", fill = "Race") +
  theme_minimal()
```
This bar chart compares the total number of stops for Black and White drivers in each city over the full time period covered by the data. It shows this by having total stops on the y-axis and cities on the x-axis. We can see two bars for each city to show how many times black people were stopped and how many times white people were stopped. While some cities show more stops for White drivers and others for Black drivers, this variation likely reflects differences in each city’s racial demographics rather than disparities in policing alone. It could also be attributed to having some incomplete or less data in some years This plot provides important context for interpreting the search rate data by showing the baseline number of stops per group.




Conclusion:
Through querying traffic stop data across six U.S. cities using SQL, I analyzed how search rates compare between Black and White individuals over multiple years. By combining city-level results into a unified dataset, I was able to observe trends within and across locations, comparing both the rate and total number of searches by race. The analysis revealed that Black drivers consistently experienced higher search rates than White drivers in every city studied. While the magnitude of disparity varied by city and year, the pattern was both persistent and systemic. Interestingly, when comparing the total number of searches, the dominant racial group varied by city—likely reflecting differences in local demographic composition rather than search behavior alone.

Citation:
Pierson, Emma, Camelia Simoiu, Jan Overgoor, Sam Corbett-Davies, Daniel Jenson, Amy Shoemaker, Vignesh Ramachandran, et al. 2020. “A Large-Scale Analysis of Racial Disparities in Police Stops Across the United States.” Nature Human Behaviour, 1–10.

Who collected the data? The data was collected and analyzed by a multidisciplinary research team led by Emma Pierson, with co-authors from Stanford University, including members of the Stanford Computational Policy Lab and other researchers from academia and civic technology.


Why? The data was gathered to conduct one of the largest empirical investigations into racial disparities in police traffic stops in the United States. The study sought to measure disparities in stop rates, search decisions, and outcomes, and to test whether these disparities could be explained by differences in behavior versus systemic bias.


Original source: The underlying stop data was obtained from publicly available datasets released by state and local law enforcement agencies across the U.S., covering over 100 million traffic stops. The Stanford Open Policing Project (https://openpolicing.stanford.edu) played a central role in assembling and standardizing this data for research use.




