---
title: "StringAnalysis"
description: |
 [Raw data source](http://www.amberboydstun.com/)
author: Nathan Dhanani
date: March 5, 2025
format: html
---

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
library(tidyverse)
library(lubridate)
library(RTextTools)
data(NYTimes)
as_tibble(NYTimes)
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
NYTimesv2 <- NYTimes |>
  mutate(politics_mention = str_detect(Title,regex("election|vote|senate|president|congress|politics|congress|clinton|trump", ignore_case = TRUE))) |>
  mutate(title_word_count = str_count(Title,"\\S+")) |>
  mutate(location = str_extract(Title, regex("New York|California|Texas|France|Russia|China|Japan|Mexico|Bosnia|Serbia|Germany|UK|USA", ignore_case = TRUE)))|>
  mutate(location = str_to_lower(location)) |>
  filter(politics_mention == TRUE, !is.na(location)) |>
  group_by(location) |>
  summarize(article_count = n(), .groups = "drop")
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
ggplot(NYTimesv2, aes(x = reorder(location, -article_count), y = article_count, fill = location)) +
  geom_bar(stat = "identity") +
  labs(title = "Most Mentioned Locations in Political Articles",
       x = "Location",
       y = "Number of Articles") +
  theme_minimal()
```

The plot displays the frequency of political articles mentioning specific locations. Each bar represents a location, with the height indicating the number of articles referencing it. Locations are ordered from most to least mentioned, providing a clear view of geographic trends in political news coverage.

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
library(lubridate)
NYTimesv3 <- NYTimes |>
  mutate(Date = dmy(Date),  
         Year = year(Date),
         politics_mention = str_detect(Title, regex("election|vote|senate|president|congress|politics|congress|clinton|trump", ignore_case = TRUE)),
         title_word_count = str_count(Title,"\\S+"),
         location = str_extract(Title, regex("New York|California|Texas|France|Russia|China|Japan|Mexico|Bosnia|Serbia|Germany|UK|USA", ignore_case = TRUE))) |>
  filter(politics_mention == TRUE, !is.na(location)) |>
  group_by(Year) |>
  summarise(article_count = n(), .groups = "drop") |>
  arrange(Year)
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
ggplot(NYTimesv3, aes(x = Year, y = article_count)) +
  geom_line(color = "blue", size = 1) +
  geom_point(size = 2, color = "red") +
  labs(title = "Political Articles Mentioning Key Locations Over Time",
       x = "Year",
       y = "Number of Articles") +
  theme_minimal()
```

This line graph illustrates the number of political articles referencing key locations across different years. The x-axis represents the year, while the y-axis shows the total number of articles that mention political topics alongside specific locations. The blue line indicates trends over time, while red points highlight the exact number of articles per year. This visualization helps identify fluctuations in political coverage across the dataset.
