---
title: "StringAnalysis"
description: |
 [Raw data source](http://www.amberboydstun.com/)
author: Nathan Dhanani
date: March 5, 2025
format: html
---

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
library(tidyverse)
library(lubridate)
library(RTextTools)
data(NYTimes)
as_tibble(NYTimes)
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
NYTimesv2 <- NYTimes |>
  mutate(politics_mention = str_detect(Title,regex("election|vote|senate|president|congress|politics|congress|clinton|trump", ignore_case = TRUE))) |>
  mutate(title_word_count = str_count(Title,"\\S+")) |>
  mutate(location = str_extract(Title, regex("New York|California|Texas|France|Russia|China|Japan|Mexico|Bosnia|Serbia|Germany|UK|USA", ignore_case = TRUE)))|>
  mutate(location = str_to_lower(location)) |>
  filter(politics_mention == TRUE, !is.na(location)) |>
  group_by(location) |>
  summarize(article_count = n(), .groups = "drop")|>
    mutate(location = fct_reorder(location, article_count))  

```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
ggplot(NYTimesv2, aes(x = location, y = article_count, fill = article_count)) +
  geom_bar(stat = "identity") +
  labs(title = "Most Mentioned Locations in Political Articles",
       x = "Location",
       y = "Number of Articles") +
  theme_minimal()
```

This bar chart shows which geographic locations appear most frequently in political articles published by *The New York Times*. Each bar represents a specific location mentioned in political headlines, based on keywords like “election,” “vote,” “president,” etc. The data has been filtered to include only articles that reference both politics and one of several predefined countries or regions. The color gradient indicates the volume of mentions, making it easier to spot which places dominate the political conversation in NYT headlines.

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
library(lubridate)
NYTimesv3 <- NYTimes |>
  mutate(Date = dmy(Date),  
         Year = year(Date),
         politics_mention = str_detect(Title, regex("election|vote|senate|president|congress|politics|congress|clinton|trump", ignore_case = TRUE)),
         title_word_count = str_count(Title,"\\S+"),
         location = str_extract(Title, regex("New York|California|Texas|France|Russia|China|Japan|Mexico|Bosnia|Serbia|Germany|UK|USA", ignore_case = TRUE))) |>
  filter(politics_mention == TRUE, !is.na(location)) |>
  group_by(Year) |>
  summarise(article_count = n(), .groups = "drop") |>
  arrange(Year)
```

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
ggplot(NYTimesv3, aes(x = Year, y = article_count)) +
  geom_line(color = "blue", size = 1) +
  geom_point(size = 2, color = "red") +
  labs(title = "Political Articles Mentioning Key Locations Over Time",
       x = "Year",
       y = "Number of Articles") +
  theme_minimal()
```

This line chart tracks the number of political articles published each year by *The New York Times* that mention at least one major global location. By focusing only on political headlines containing specific keywords and country names, this plot visualizes how global political focus in NYT coverage has shifted or intensified over time. Spikes or dips in the line may correspond to key events such as presidential elections, international conflicts, or geopolitical crises.
